\chapter{Acknowledgements}
\enlargethispage{\baselineskip}

Most of all, I would like to express my deepest gratitude to my supervisor Professor Monica Divitini. You have always been present and supportive for me. Under your guidance I grew up as a researcher and as a person. Thanks for the patience and the effort you dedicated to me.

I am extremely grateful to my co-advisors Professor John Krogstie and Dr Simone Mora. You have been an example of integrity and professionalism, I was inspired by your hard work and always found support and advice when in need.

I wish to thank all the colleagues and co-authors in our research group who contributed to my work: Anna, Lisa, Mikael and Dag Rune.

I extend my gratitude to the colleagues from the Department of Computer Science and the ISSE group, especially to my mates Elena, Stefano, Massimiliano, Katerina and Sofia for the engaging and constructive discussions.

Thanks to all my friends in Norway and in Italy for always being supportive, curious and creative, you inspired me to strive for more.

Thanks to my family: Valeria, Massimo, Matteo, Daniele, and to my grandparents Piero, Piera, Cesi and Ugo. Without you all of this would not have been possible, each of you built a piece of who I am today.

A big thank you to Cesilie, who had the rare privilege of being the one to stand \textit{grumpy}-me, \textit{late-working}-me and \textit{ranting}-me. Thanks for patiently listening, for always having helpful words and for the hugs of encouragement.

\begin{flushright}
\emph{\small Francesco Gianni}\\\emph{\small \today}
% February 26th, 2019}	
\end{flushright}
